---
title: "Simple Descriptive Techniques for Time Series Analysis"
author: "Amir R"
output:
  html_document:
    df_print: paged
---
```{r Packages, eval=T, include=FALSE}
library(magrittr) #for the pipes
# install.packages("magrittr") if not installed
```

#Objectives




***


Let's load some sample data
```{r}
load("TSsimple.RData")
```

and examine its structure

```{r}
blowfly<-read.csv("blowfly.csv", header=T) #delete later
str(blowfly)
```
***
Time for a first look at Nicholson's flies 
```{r}
plot(blowfly$total)
summary(blowfly$total)
```

and using a time series plotting method:
```{r}
plot.ts(blowfly$total)
```

#The `ts` class

`plot.ts` plots the object as if it were of the special class of R objects known simply as *ts*

the `ts()` function is used to convert objects to the ts class

```{r}
flies<-ts(blowfly$total)
class(flies)
typeof(flies)
```
many base/stats functions are actually designed for the 'ts' class
example:`acf`
```{r}
plot(flies)
acf(flies)
y.filter <- filter(flies, filter = rep(1, 5))/5
plot(y.filter)

deltat(flies)

plot(window(flies,100,200,1))

```
but using  them on a non- *ts* object may or may not result in an error:
```{r eval=FALSE, include=FALSE}
acf(blowfly$total)

y.filter <- filter(blowfly$total, filter = rep(1, 5))/5
plot(y.filter)

deltat(blowfly$total)
frequency(blowfly$total)
```

Warning: the `ts` class is designed only for equidistant series. other packages have other classes and functions which extend capabilities

***
#descriptive techniques and time series exploration

let's return to our example
```{r echo=FALSE}
plot(flies)
```

This is an example of a relatively clear cyclical process, at least at the begining (points 1:200). the second part (200:361) looks less regular and with an upward trend.

Exploratory Questions

 * what are the cycle periods ?
 * is there any hidden periodical behaviour that is not immediately apparent ?
 * Do these 2 parts behave similarly or is there a fundamental change ? which attributes  of the first part are conserved in the second part
 * can we tie this behaviour to external parameters ?

Other, not so clear examples:
```{r}

```

We can see some clearer cyclical behaviour of a section if we zoom in and explore a bit with `window()`:

```{r}
plot(window(flies,1,100))
plot(window(flies,1,100,deltat = 5))
plot(window(flies,1,100,deltat = 10))
plot(window(flies,1,100,deltat = 20))
```
We finally lose the cyclical behaviour at ~detat=20. let's remember this...



The flies series is unusually clear, and still - its cyclical behaviour is only one possible pattern.

other series may represent processes that behave in a way that is not immediately recognizable, and have many things going at once.
example:
```{r}

```


We naturally want  tools that will describe the entire series' and parts of the series' attributes, be exectuable fast in R, and guide us in further steps (such as model fitting and forecasting). 

for that, we need to undestand autocorrelation and related attributes of time series

##Measures of dependence 

###Background on Autocorrelation

how does my data behave in relation to itself ? to illustrate the background for autocorrelation:

let's look at a lag of 1 point, or 'how does this week's population relate to last week's population ?

```{r}
plot(flies[1:350],flies[2:351])
```

Clearly, there's a pattern at lag =1, at least the smaller population sizes

now let's look at other lags:

```{r fig.show='hold'}
#lag=2
plot(flies[1:350],flies[3:352])
#lag=3
plot(flies[1:350],flies[4:353])
```

we start losing the pattern gradually...

more lags, lag = 1 to lag =8:

```{r echo=FALSE, results='hide'}
par(mfrow=c(2,4))
sapply(1:8,function (x) plot(flies[-c(1:x)],flies[-c(361:(361-x+1))]))
```

however, at lags 9-16:

```{r echo=FALSE, message=FALSE, results='hide'}
par(mfrow=c(2,4))
sapply(9:16,function (x) plot( flies[-c(1:x)],flies[-c(361:(361-x+1))]))
par(mfrow=c(1,1))
```

**summary**

 * we begin with a pattern, lose it, and then find it again
 * at smaller populations, we begin with a positive correlation, lose it, then a negative correlation  which peaks at about lag = 10, lose it, and so the cycle goes.


All of this is quite cumbersome !

### The Correlogram, using `acf()`
```{r}
acf(flies)
```

This summarizes and charts the autocorrelation structure in the series, up to a default lag size, which is 25 in this case.

The vertical axis is the relative magnitude of the  **Sample Autocorrelation Coefficients**:
let's have a look at them


```{r results='hide'}
a<-acf(flies, plot = F)
```

```{r }
a$acf %>% head(10) #for the actual factors
```

and the horizontal axis is simply the lag size.

```{r echo = FALSE}
acf(flies)
```

there's no better evidence of cycles, the flies exhibit strong positive AND negative correlation at regular intervals throughout the series, with a cycle period of 19 weeks.



let's look at the entire series
```{r}
acf(flies, lag.max = 360,col="red")
```

We can see

* the relative power of correlation at different lags
* the (sort of regular) cyclical pattern, espescially at the beginning
* the dimininshing positive correlation
* some alternation between positve and negative correlations, which is typical of cycles.

warning: there is bias here with incresing lag size. why ?

it's fair to conclude, without further evidnece to the contrary, that perhaps

  * there's some kind of a diminishing but long term "memory" in the system, up to a lag of 100 weeks. events that happened 50 weeks ago still have an influence.
  * clear cycles with a period of 19 weeks
  * there's a **LOT** of non-randomness in the sequence

This last point is espesically important for linear and other modeling, which assume that samples are independent. we may want to stop and think before going on to perform a conventional `lm()` ! , and at the very least be aware of series' behaviour.

and this illustrates why we need to explore time series intenrally before we go on to the other, external parameters.

```{r echo=FALSE, fig.show='hold'}
par(mfrow=c(2,1))

plot(flies,main=NULL)
acf(flies, lag.max = 360, main=NULL)

par(mfrow=c(1,1))
```

The **Sample Autocovariance Coeefficients**

We could also look at the correspoding covariance coefficients by change the type parameter.
```{r}
acf(flies,type="covariance")$acf %>% head(10)
```


***
###Partial Autocorrelation
Correlograms produced using the sample *autocorrelation* coefficients do not account for the fact that for a given lag size there may be correlation between internal points, (eg. values that are 4 points apart were correlated, but so do values that are 2 points apart).

We sometimes want to control for the internal correlations inside the lag, or in other words, to check what would have been the correlation coefficients, had all the internal lags' coefficients were forced to zero. 

the **Partial Autocorrelation ** is the relationship between this week's population and the population at lag n when we have already controlled for the correlations between all od the successive weeks between this week and week n

We only need to use `pacf` or `acf(type="partial")` instead of the default

```{r}
#Autocorrelation Coefficients
acc<-a$acf %>% head(12) %>% round(2)
#Partial Autocorrelation Coefficients
partial.acc<-pacf(flies, plot = F) %>% .$acf %>% head(12) %>% round(2)

plot(1:12,acc,xlab = "Lag",ylab="Coefficient")
points(1:12,partial.acc,col = "green")
lines(1:12,rep(0,12))
legend("topright", c("ac coefficients", "Partial ac coeffients"),col = c("black","green"),pch = c(1,1))

pacf(flies)
pacf(flies, lag.max = 360,col="red")
```
what can we take from this ?
the maximal negative and maximal positive lags are important places in the cycle.
clues for possible underlying processes in lags 3 and 12

the PACF can be thought of as analogous to a derivative - points driving change in pitch or direction. 


***

Summary till now, using just Correlograms for the entire series
* cycle length - about 19
* Lags 2-4, 12-16 are suspected, we should look for biological mechanisms there, when we construct an explanatory model.

again, this is only from the time series itself, before considering any other data.


###Cross Correlation
The total number of flys is not the only one we have...

plotting the dataframe variables against eachother
```{r}
plot(blowfly)

```

we get a strong correlation between total, deaths, and emerging. but there's an inherent problem in this analysis...

```{r, fig.show='hold'}
blowfly %>% ts ->bf
plot(bf)


blowfly$eggs %>% ts ->eggs
blowfly$nonemerging %>% ts ->nonemerging
blowfly$emerging %>% ts ->emerging

ccf(flies,eggs)
ccf(flies,nonemerging)
ccf(flies,emerging)
```
Reminder: Lags 2-4 (negative), 12-16 (positive) are suspected from `pacf()` of the total of "flies"

from `ccf()` and `pacf()`

`pacf`:
eggs - negative at lag 5
non emerging - negative at lag 3
emerging - negative at lag 4

`ccf`:

eggs vs. total- negative at lags 2-3 and (-17 - -16), positive at 12-13
non emerging vs. total - negative at lag 3
emerging vs. total - negative at lag 4

Note: positive and negative lags are non-symmetric when in comes to correlation coefficients.


findings:
cycle length = still 19

larval processes *appear* to drive the cycles


***
##analysing parts of a TS

now let's adapt the analysis for the two parts, separately


```{r, fig.show='hold'}
first<-ts(flies[1:200])
second<-ts(flies[201:361])
par(mfrow=c(1,2))
plot(first, ylim = c(0,14000))
plot(second)
par(mfrow=c(1,1))

```

***

1.visually identify a breakpoint

2. apply the same steps on parts of the series to find their attributes and identify discontinuities
```{r acf on both parts, fig.show="hold"}
par(mfrow=c(2,2))
acf(first)
acf(second)
pacf(first)
pacf(second)
par(mfrow=c(1,1))
```
more or less the same attributes

 * same shapes
 * cycle lengths is still 19 weeks
 * Important Lags in the second parts are shifted 1 or -1 lags , compared with the first - change points are closer to each other
 
 however - 


```{r}

```

##Smoothing


## Quick breakdown


## Conclusion & Remarks


# Source Material and useful links


