---
title: "lubridate practice"
author: "Tal G"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(hms)
library(scales)
library(knitr)
theme_set(theme_bw())
```

```{r}
timeSample <- tibble(
  date_1 = c("02-10-2019", "2/10/19"),
  date_2 = c("2019 10 02", "19/10/2"),
  datetime_1 = c("02-10-2019 19:00:00", "02-10-2019 7:00:00PM")
)
```

```{r, echo=FALSE, results='asis'}
knitr::kable(timeSample)
```

# Part 1 - simple example

### Convert to Datetime object (class POSIXCT/POSIXIT)

`lubridate` has multiple functions that convert factor (or any other class) to dates. For example:

*  `ymd()` (year, month, day)
*  `mdy()`
*  `dmy()`

The difference is the order of the time variables--you need to choose the one that fits for your date type.

e.g. in the `date_1` column the date is built in the day-month-year format so I use the dmy function

```{r}
timeSample %>% 
  mutate(date_1 = dmy(date_1)) %>% 
  kable()
```

in date_2 column the date is built in format of year-month-day so i use the ymd function

```{r}
timeSample %>% 
  mutate(date_1 = dmy(date_1),
         date_2 = ymd(date_2)) %>% 
  kable()
```

**Note** the only thing that matters is the order of the day/month/year. The separators can change (':' or '/' or '-' etc.) and lubridate can handle it...the same goes for 02 or 2, 2019 or 19. 

** Really cool!! **

`lubridate` can also handle date and time object in the same manner...
* note - can translate AM/PM to the right 24 clock hour

```{r}
timeSample <- timeSample %>% 
  mutate(date_1 = dmy(date_1),
         date_2 = ymd(date_2),
         datetime_1 = dmy_hms(datetime_1)) 
```


```{r, echo=FALSE, results='asis'}
kable(timeSample)
```

### Time zones

When dealing with times in R it's super important to pay attention to time zones

If you don't specify the time zone, the default will be UTC. This can cause problems later on if you do all kind of time calculations and don't understand why you get weird numbers (;

The `force_tz` function will take the time as it and change the time zone (in the 'back office') to the required one. It's good for cases where you didn't specify the time zone for your data beforehand and now you want to fix it.

```{r}
tz(timeSample$datetime_1)

timeSample$datetime_1 <- force_tz(timeSample$datetime_1, "Asia/Jerusalem")

tz(timeSample$datetime_1)
```

The `with_tz` function will **convert** your time zone to a different one.
this function is good for cases where you want to know the time in different places

```{r}
timeSample$datetime_utc <- with_tz(timeSample$datetime_1, "UTC")
```

```{r echo = FALSE}
kable(timeSample)
```

** The best practice is to specify the timezone in first place **

```{r}
ymd_hms("2019-07-20 08:35:00", tz = "Asia/Jerusalem")
```


### Other nice functions in `lubridate`

```{r}
Sys.time() # will give the current date and time

days_in_month(today()) # how many days in this month?
days_in_month(today() + 31) # how many day in the month 31 days from now?

per <- period(hours = 10, minutes = 5) # define a period you can add to other date time...

per # a period class object

Sys.time() + per

Sys.time() - per

difftime(Sys.time() + per, Sys.time(), unit = "hours")

difftime(Sys.time() + period(week = 3), Sys.time(), unit = "day")

quarter(Sys.time(), with_year = T) # the qurater of the year...
```

## Part 2  - Taxi data frame

```{r eval=FALSE, include=FALSE}
# taxi <- read_csv("chicago-taxi-rides-2016/chicago_taxi_trips_2016_01.csv")
# taxi <- taxi[1:5000, ]
# colnames(taxi)
# taxi <- select(taxi, taxi_id, trip_start_timestamp, trip_end_timestamp, trip_miles, fare, tips, tolls, trip_total, payment_type, company)
# 
# colnames(taxi) <- c("taxi_id", "trip_start", "trip_end", "trip_miles", "fare", "tips", "tolls", "trip_total", "payment_type", "company")
# 
# write.csv(taxi, "taxi.csv", row.names = FALSE)
```

Import the data

```{r}
taxi <- read_csv("taxi.csv")
```

Convert to time class ("POSIXct"  or "POSIXt" )

```{r}
class(taxi$trip_start)
class(taxi$trip_end)

taxi <- taxi %>% 
  mutate(trip_start = ymd_hms(trip_start, tz = "America/Chicago"),
         trip_end = ymd_hms(trip_end, tz = "America/Chicago"))
```


### What is the week-day with the highest number of taxi trips?

`lubridate` has a series of functions that extract different variables out of the datetime

* `day()`
* `month()`
* `year()`
* `week()`
* `hour()`
* etc.

```{r}
# If label is false then is will show the week day numbers (1:7)
wday(taxi$trip_start, label = TRUE) %>% head()
```

```{r}
taxi %>% 
  mutate(week_day = wday(trip_start, label = TRUE)) %>% 
  ggplot(aes(week_day)) +
  geom_bar()
```

Friday is the most busy day...

### What is the duration of each taxi trip?

`?difftime`

```{r}
taxi <- taxi %>% 
  mutate(duration = difftime(trip_end, trip_start, 
                             units = "mins")) %>% 
  filter(duration < 100) #remove the abnormally long rides
  
taxi %>% 
  ggplot(aes(duration)) +
  geom_histogram(bins = 15) +
  xlim(-5, 100)
```

### It seems that trips with duration below 15 mins default to zero...

Let's assume that this is a mistake and change all the zero duration to a five minus drive

```{r}
five_min <- as.period(5, unit = "mins") # create five minute object

taxi <- taxi %>%
  
  # if the duration is 0 add five minutes
  mutate(alt_end_time = ifelse(duration == 0,
                               trip_end + five_min,
                               trip_end)) %>% 
  # the output is in numeric format - change to datetime format and don't forget the time zone
  mutate(alt_end_time = as_datetime(alt_end_time, tz = "America/Chicago")) %>% 
  
  # fix the duration
  mutate(fix_duration = difftime(alt_end_time, trip_start, 
                                 units = "mins"))

taxi %>% 
  filter(fix_duration < 100) %>% 
  ggplot(aes(fix_duration)) +
  geom_histogram(bins = 15) +
  xlim(-5, 100)
```

# When are the longest trips? Morning, noon or night?

```{r}
taxi <- taxi %>% 
  mutate(hour = hour(trip_start),
         time_category = case_when(
           between(hour, 5, 12) ~ "morning",
           between(hour, 12, 18) ~ "noon",
           TRUE ~ "night"
         )) 

ggplot(taxi, aes(time_category, trip_miles)) +
  geom_boxplot() +
  scale_y_log10() +
  coord_flip()
```

# Round dates and times

Assume I want to round the trips to half hour unit so I can group them better...

the unit can be other combinations as well...

e.g.

*  `"2 day"`
*  `"5 mins"`
*  `"quarter"`

```{r}
(taxi <- taxi %>% 
  mutate(trip_start = round_date(trip_start, 
                                 unit = "1 hour")))
```

Other functions are `floor_date()` and `ceiling_date()` that will round the time down or up respectively

### Plots:

A basic plot

```{r}
(p <- ggplot(taxi, aes(trip_start, fare)) +
  geom_point(color = "blue", alpha = 0.1) + 
  xlab("Trip start") + ylab("Trip cost ($)"))
```

To control the time axis use `scale_x_datetime()` or `scale_x_date()` if you only have date object

```{r}
p + scale_x_datetime(date_breaks = "5 day", 
                     labels = date_format("%d-%m-%Y"))
```

The `date_break` argument controls for the tick mark

The `labels` argument controls for the label of your data

Note that the syntax here is different...

before each time variable, you need to add % sign, then the symbol for the time argument and the desired separator

link to the date symbol guide:

link[]

```{r}
p + scale_x_datetime(date_breaks = "12 hours",
                     labels = date_format("%d-%m-%y %H:%M")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

How to show only some of the time series

```{r}
start_time <- ymd_hms("2016-01-01 06:00:00", tz = "Asia/Jerusalem")
end_time <- ymd_hms("2016-01-05 06:00:00", tz = "Asia/Jerusalem")

# create a start and end time R object
start_end <- c(start_time, end_time)

p + scale_x_datetime(limits = start_end,
                     date_breaks = "12 hours", 
                     labels = date_format("%d-%m-%y %H:%M")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## By hour plots

If you want to plot hourly patterns  there are 2 options

#### 1. use the `hms` package that can extract only the time variable from the date-time

```{r}
taxi <- taxi %>% 
  mutate(time = as.hms(trip_start)) 

ggplot(taxi, aes(time, fare)) +
  geom_point(color = "blue", alpha = 0.1) +
  scale_x_time(labels = function(x) strftime(x, "%H:%M"))
```

#### Option 2

Give one fake date to all the observation and customize with the `scale_x_datetime()` arguments

```{r}
taxi <- taxi %>% 
  mutate(fake_date = ymd_hms(paste("2000-01-01", time)))

(p_2 <- ggplot(taxi, aes(fake_date, fare)) +
    geom_point(color = "blue", alpha = 0.1) + 
    xlab("Trip start (hour)") + 
    ylab("Trip cost ($)"))

p_2 + 
  scale_x_datetime(date_breaks = "2 hours", 
                   labels = date_format("%H:%M")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



